---
title: "Packageless coding baseR"
author: "Ali Martin"
date: "11/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Creating a vector 

```{r}
#numeric vector
my_vector <- vector(mode = "numeric", length = 100)

#logical vector
my_second_vector <- vector(mode = "logical", length = 20)
```

## vector Properties 

### Length

```{r}
#length of my vector
length(my_vector)
length(my_second_vector)
```

### Class

```{r}
#what's in my vector?
class(my_vector)
class(my_second_vector)
```

### Indexing (accessing) vectors

```{r}
# using c to create a vector
my_vector <- c(1:50)

# what's the first element? 
my_vector[1]

# what's the last element?
my_vector[-1] #pulls everything else but first element

# what's the seventh element
my_vector[7]

# select a range from a vector
my_vector[3:7]
```

## Matrices

### creating matrices

```{r}
# create a 10 x10 matrix full of NA's

my_matrix <- matrix(data = 1:100, nrow = 20, ncol = 5)
```

### matrix properties

### dimensions

```{r}
# what are the dimensions?
dim(my_matrix) #number of rows is always first, second is number of cols
```

### Length

```{r}
# how many elements are there?
length(my_matrix) #there are 100 elements
```

#### Class

```{r}
class(my_matrix)
```

### Indexing

```{r}
my_matrix[2,] # extract values from 2nd row

# index a column
my_matrix[,3] #extract values from 3rd column

# index a given cell
my_matrix[8,3] 

# index and replace
my_matrix[8,3] <- 100000

# replace an entire column
my_matrix[,5] <- 10

```

# Control flow

## `for` loops 

variable (counter), sequence, expression

```{r}
for (i in 1:10)( #counter is i, sequence is 1:10
  print ('Hello world') #expression
)
```

# generating random numbers

```{r}
rnorm(n = 5, mean = 10, sd = 2) #rnorm is generating random numbers
```

# Let's combine all of the above

we will use `numeric` to create an empty vector
we will use `for` to iterate along items of the empty vector
we will use `[]` to access each slot and fill in that empty vector
we will use `rnorm` to sample values from a random normal distribution
We will plot the trend




## writing the code down


```{r}
# set up model parameters
K <- 100
nsteps <- 100
mu <- 0.4
sigma <- 1

#set up model objects

N <- vector(mode = "numeric", length = nsteps)
N[ 1] <- 25

# model it
for(t in 2:nsteps){
r <- rnorm(n = 1, mean = mu, sd = sigma)
N[t] <- N[t-1] + r* (1-(N[t-1] / K))
} # write expression in squiggly brackets


# Visualize it
plot(N)
```

# Something more
We will use `matrix` to create an empty matrix (rows are time, columns are runs) We will repeat the process

```{r}
# set up model parameters

K <- 100
nsteps <- 100
mu <- 0.4
sigma <- 1

#set up model objects

N <- matrix(data = NA, nrow = nsteps, ncol = 100)
N[ 1,] <- 25

# model it

for(t in 2:nsteps){
r <- rnorm(n = 100 , mean = mu, sd = sigma)
N[t, ] <- N[t-1 , ] + r * ( 1 - (N[t-1,] / K))
} # write expression in squiggly brackets


# Visualize it (plot matrices using matplot)

matplot(N, type = "l") # type linear
```

# Bring in your friends!

convert it to things we know about: `ggplot2`

```{r}
# load tidyverse
library(tidyverse)

#create data frame
my_data <- as_tibble(N) %>% 
  mutate(time = 1:nsteps) %>% 
  gather(run, N,-time )

#gather takes all observations in columns to combine into run, we want it to leave out time

#visualize it
ggplot(data = my_data, mapping = aes(x = time, y = N)) +
  geom_line(aes(group = run), size = 0.1, alpha = 0.5) +
  stat_summary(geom = "line", fun = mean, color = "red", size = 1) +
  theme_bw()+
  labs(x = "Time (years", y = "Population size (N)")
```

